Time and space specifications for stencils

### 2016-06-20 paper.tex
section on equations and inequations

@@

\bibliographystyle{plain}

@@

In some cases, it is useful to give a lower and/or upper bound for a
stencil. This can be done using either the \term{atMost} or
\term{atLeast} modifiers. This is particularly useful in situations
where there is a non-contiguous stencil, which does not fit the rest
of the specification syntax.  For example:
%%
\begin{ExmVerbatim}
!= stencil atMost, forward(depth=2, dim=1) :: a
b(i) = a(i) + a(i+2) 
\end{ExmVerbatim}
%%

\subsubsection{Temporal specifications}

@@

Figure~\ref{fig:equations}

\begin{figure}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
\framebox{$\textit{region} \equiv \textit{region}'$} \\[1em]
\begin{array}{rll}
(\textsc{F\; \texttt{+} \;F}) \;\; &
\stenFwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenFwdS{m}{d} \\[1em]
%
(\textsc{B\; \texttt{+} \;B}) \;\; &
\stenBwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenBwdS{n}{d} \; \texttt{+} \; \stenBwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;C}) \;\; &
\stenCenS{n \, \textsf{max} \, m}{d} \\
\equiv \; & \stenCenS{n}{d} \; \texttt{+} \; \stenCenS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;F}) \;\; & \stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenFwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;B}) \;\; & 
\stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenBwdS{m}{d} \\[1em]
(\textsc{B\; \texttt{+} \;F}) \;\; & 
\stenCenS{n}{d} \\
\equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenBwdS{n}{d} \\[1em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.5em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.5em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.5em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.5em] 
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T) \\[0.5em] 
(\textsc{inter}) \;\; & (S \; \texttt{+} \; S') \; \texttt{*} \;
                              (T \; \texttt{+} \; T') \\
 \equiv \; & (S \;  \texttt{*} \; S) 
 \; \texttt{+} \; (S \; \texttt{*} \; T') 
 \; \texttt{+} \; (S' \; \texttt{*} \; T)
 \; \texttt{+} \; (S' \; \texttt{*} T')  \\[0.5em] 
\end{array} \\ \\
%
\framebox{$\overline{\textit{specDec} :: \texttt{v}} \equiv 
\overline{\textit{specDec'} :: \texttt{v}}$} \\[1em]
\begin{array}{rl}
(\textsc{mutual}) \; & 
\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
  :: \texttt{w} 
\\
\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
  \texttt{v} \\[0.5em]
(\textsc{coalesce}) \; & 
\texttt{stencil} \; \texttt{dependency} (\bar{v}) 
  :: \texttt{v} \\
& ; \texttt{stencil} \; \texttt{dependency} (\bar{w}) 
  :: \texttt{v} 
\\
\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
  \texttt{v} \\[0.5em]
(\textsc{spatial}) \; &
(\textit{region} \equiv \textit{region'}) \\
\Rightarrow \; & (\texttt{stencil} \; [\textit{mod}] \; [\textit{approxMod},] \;
\textit{region} \\
& \!\!\equiv \; \texttt{stencil} \; [\textit{mod}] \;
            [\textit{approxMod},] \;
\textit{region}') \\
\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications}
\label{fig:equations}
\end{figure}

\subsubsection{Safety and sub-specifications}

@@

\begin{figure}[t]
{\framebox{
\begin{minipage}{0.88\linewidth}
\begin{align*}
\hspace{-0.7em}
\begin{array}{c}
\dfrac{}{S <: S}(\textsc{refl}) \qquad \dfrac{R <: S \quad S <: T}{R <:
  T}(\textsc{trans}) \\[1.5em]
\dfrac{}{S <: \; \texttt{unspecified dims=}ds}(\top)
\\[1.5em]
\dfrac{}{\texttt{read-once} \, S <: S}(\textsc{rep})
\\[1.5em]
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} ds \subseteq es \; \wedge \; n \leq m \hspace{3em}}
{\begin{array}{rl}
\texttt{fwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{fwd} \;
  \texttt{depth=}m \; \texttt{dim=}es
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \texttt{bwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{bwd} \;
  \texttt{depth=}m \; \texttt{dim=}es
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \texttt{fwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{sym} \;
  \texttt{depth=}m \; \texttt{dim=}es
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \texttt{bwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{sym} \;
  \texttt{depth=}m \; \texttt{dim=}es
\end{array}}
%{\footnotesize{(\textsc{cover})}}
%\hspace{-0.1em}
\end{array}
\end{align*}
\end{minipage}}}
\caption{Definition of sub-specification relation}
\label{fig:inequations}
\end{figure}

@@

\section{Examples and common patterns}

% Implicit vs explicit differentiation methods
% (forward implicit needs 'solver')
% backward - Euler
% runge-kutta 2nd-order

\begin{example}[FTCS (Forward time, centered space)]

  Otherwise known as the \emph{explicit method}, is a common
  pattern since it tends to be more efficient than other approches
  (see BTCS and CTCS below). However, its disadvantage is that it is
  unstable if the resolution is not well chosen.

\begin{ExmVerbatim}
b(x) = a(x) + r*(a(x-1) - 2*a(x) + a(x+1))
\end{ExmVerbatim}
%
For this statement, the inference provides the specification:
%
\begin{SpecVerbatim}
!=  stencil centered, depth=1, dim=0 :: a
\end{SpecVerbatim}

\end{example}

\paragraph{BTCS (Backward time, centered space) -- implicit method}

\paragraph{CTCS (Centered time, centered space) - Crank-Nicolson method}

>>>>

\definecolor{darkblue}{rgb}{0.0,0.0,0.5}
\definecolor{darkgreen}{rgb}{0.0,0.4,0.0}

\usepackage[]{hyperref}
\hypersetup{
    unicode=false,          % non-Latin characters in Acrobat's bookmarks
    pdftoolbar=true,        % show Acrobat toolbar?
    pdfmenubar=true,        % show Acrobat menu?
    pdffitwindow=false,      % page fit to window when opened
    pdftitle={},    % title
    pdfauthor={}
    pdfsubject={},   % subject of the document
    pdfnewwindow=true,      % links in new window
    pdfkeywords={keywords}, % list of keywords
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=darkblue,          % color of internal links
    citecolor=darkblue,        % color of links to bibliography
    filecolor=green,      % color of file links
    urlcolor=blue,          % color of external links
}

@@

\bibliographystyle{abbrvnat}

@@

In some cases, it is useful to give a lower and/or upper bound for a
stencil. This can be done using either the \term{atMost} or
\term{atLeast} modifiers. This is particularly useful in situations
where there is a non-contiguous stencil, which does not fit the rest
of the specification syntax.  For example:
%%
\begin{ExmVerbatim}
!= stencil atLeast, reflexive(dims=1)      :: a
!= stencil atMost, forward(depth=2, dim=1) :: a
b(i) = a(i) + a(i+2) 
\end{ExmVerbatim}
%%
\subsubsection{Temporal specifications}

\subsection{Note on the design}

The names ``forward'', ``backward'' and ``centered''
are inspired by standard terminology in numerical analysis
for the shape of discretisation schemes. For example,
the standard ``explicit method'' for approximation
PDEs is the \emph{Forward Time, Centered Space} (FTCS)
scheme~\cite{dawson1991finite}. For
the one-dimensional heat equations, an FTCS discretisation 
provides approximation code with the following stencil~\cite{recktenwald2004finite}:
%%
\begin{ExmVerbatim}
do i=2:n-1
  u(i) = r*v(i-1) + r2*v(i) + r*v(i+1)
end do
\end{ExmVerbatim}
%%
where \texttt{r} and \texttt{r2} are some constants. 
A valid specification for this stencil is then:
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1) 
         + centered(depth=1, dim=2) :: v
\end{ExmVerbatim}
%%
Such a specification can be inferred, or can be inserted by a user
and checked against the code.

@@

\begin{figure}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
\framebox{$\textit{region} \equiv \textit{region}'$} \\[1em]
\begin{array}{rll}
(\textsc{F\; \texttt{+} \;F}) \;\; &
\stenFwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenFwdS{m}{d} \\[1em]
%
(\textsc{B\; \texttt{+} \;B}) \;\; &
\stenBwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenBwdS{n}{d} \; \texttt{+} \; \stenBwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;C}) \;\; &
\stenCenS{n \, \textsf{max} \, m}{d} \\
\equiv \; & \stenCenS{n}{d} \; \texttt{+} \; \stenCenS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;F}) \;\; & \stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenFwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;B}) \;\; & 
\stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenBwdS{m}{d} \\[1em]
(\textsc{B\; \texttt{+} \;F}) \;\; & 
\stenCenS{n}{d} \\
\equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenBwdS{n}{d} \\[1em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.5em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.5em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.5em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.5em] 
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T) \\[0.5em] 
(\textsc{inter}) \;\; & (S \; \texttt{+} \; S') \; \texttt{*} \;
                              (T \; \texttt{+} \; T') \\
 \equiv \; & (S \;  \texttt{*} \; S) 
 \; \texttt{+} \; (S \; \texttt{*} \; T') 
 \; \texttt{+} \; (S' \; \texttt{*} \; T)
 \; \texttt{+} \; (S' \; \texttt{*} T')  \\[0.5em] 
\end{array} \\ \\
%
\framebox{${\textit{specDec} :: \texttt{v}} \equiv 
{\textit{specDec'} :: \texttt{v}}$} \\[1em]
\begin{array}{rl}
(\textsc{mutual}) \; & 
\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
  :: \texttt{w} 
\\
\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; & 
%\texttt{stencil} \; \texttt{dependency} (\bar{v}) 
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w}) 
%  :: \texttt{v} 
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
(\textsc{spatial}) \; &
(\textit{region} \equiv \textit{region'}) \\
\Rightarrow \; & (\texttt{stencil} \; [\textit{mod}] \; [\textit{approxMod},] \;
\textit{region} \\
& \!\!\equiv \; \texttt{stencil} \; [\textit{mod}] \;
            [\textit{approxMod},] \;
\textit{region}') \\
\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications}
\label{fig:equations}
\end{figure}

\paragraph{Equations}
Figure~\ref{fig:equations} lists the equational theory for
specifications. This is broken down into equations on regions only,
given by the equivalence relation $\textit{region} \equiv \textit{region}'$ and on
on specification declarations assigned to variables via the equivalence relation
$\textit{specDec} :: \texttt{v} \equiv
\textit{specDec'} :: \texttt{v}$.

To save space we use abbreviations 
\term{fwd}, \term{bwd}, and \term{cen} for \term{forward},
\term{backward}, and \term{centered} regions respectively.  The
equations reveal something of their semantics.  The first six
equations (with labels of the form $\ldots \texttt{+} \ldots$) 
explain the behaviour of overlapping regions with sum
\term{+}. The next six equations explain the algebraic behaviour of
\term{+} and \term{*}.  Together, we see that \term{+} is an
associative, commutative, idempotent binary operator that distributes
with \term{*}, which is associative and commutative. This distribution
of \term{*} over \term{+} is key as it is used internally to give a
normal form for stencil specifications akin to Disjunctive Normal Form
(DNF) (where \term{+} is taken as ``disjunction'' and \term{*} as the
``conjunction''). We revisit this normalisation in Section~\ref{}.

\begin{figure}[t]
\begin{minipage}{0.88\linewidth}
\begin{align*}
\hspace{-0.7em}
\begin{array}{c}
\dfrac{}{S <: S}(\textsc{refl}) \qquad \dfrac{R <: S \quad S <: T}{R <:
  T}(\textsc{trans}) \\[1.5em]
\\[1.5em]
\dfrac{}{\texttt{read-once} \, S <: S}(\textsc{rep})
\\[1.5em]
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} m \leq n \hspace{3em}}
{\begin{array}{rl}
\stenFwdS{m}{d} & \stenFwdS{n}{d}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \stenBwdS{m}{d} & <: \stenBwdS{n}{d}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenCenS{m}{d} & <: \stenCenS{n}{d}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdS{m}{d} & <: \stenCenS{n}{d} \\
\wedge \; \stenFwdS{m}{d} & <: \stenCenS{n}{d} \\
\end{array}}
%{\footnotesize{(\textsc{cover})}}
%\hspace{-0.1em}
\end{array}
\end{align*}
\end{minipage}
\caption{Inequations on specifications}
\label{fig:inequations}
\end{figure}

\paragraph{Inequations: sub-specifications}

@@

@@
